Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-11 paper/paper.tex

{\color {Myblue}Rewrite at the end: 
Section~\ref{sec:defs} introduces the syntax and semantics of the
language used to illustrate our
analysis.
% Section~\ref{sec:operational} describes the operational semantics of
% the language.
Liveness analysis for this language is  described in
Section~\ref{sec:liveness}. This section also has by  a  sketch  of  a
correctness proof  relative  to  a  non-standard
semantics.  Section~\ref{sec:computing}  shows how to  encode liveness
as   finite-state  automata.    Section~\ref{sec:experiments}  reports
experimental  results  and Section~\ref{sec:lgc-always-better}  proves
that a liveness based collector  can never do more garbage collections
than a reachability based collector.
% and Section~\ref{sec:conclusion} concludes the paper.
}
\section{The target language---syntax and semantics}
\label{sec:defs}
Figure~\ref{fig:lang-syntax} describes the syntax  of our language. It
is a first order language with lazy semantics. Programs are restricted
to        be        in        Administrative        Normal        Form
(ANF)~\cite{chakravarty03perspective} where  all actual  parameters to
functions  are  variables.  While  this  restriction  does not  affect
expressibility,  this form  has  the benefit  of  making explicit  the
creation of  closures through  the $\LET$  construct.  $\LET$s  in our
language  are lazy;  in the  expression $\LET\,\,  x \leftarrow  s\,\,
\IN\,\, e$, $x$ may occur in  $s$. This enables creation of graph-like
structures in  a pure language. The  restriction of \LET\ to  a single
definition  is for  ease  of  exposition---generalization to  multiple
definitions does not add conceptual difficulties.  We further restrict
each variable in a program to  be distinct, so that no scope shadowing
occurs---this  simplifies  the  reasoning   about  the  program.   The
$\SRETURN$  and   \SIF\  expressions  and  strict   operators  trigger
evaluation of closures.

@@

\section{Liveness}\label{sec:liveness}

@@

\section{The Garbage Collection Scheme}

Our experimental setup consists of  an interpreter for our language, a
liveness  analyzer, and  a single  generation copying  collector.  The
garbage collector  can be  easily configured to  work on the  basis of
reachability (RGC  mode) or use  the liveness DFAs (LGC  mode).  While
exploring the  activation record of a  function, LGC uses  the DFAs at
the latest program  point traversed in the function's  body.  Thus, if
LGC is invoked  at the program point $\pi$, it  would use the liveness
DFAs  at  $\pi$ itself  for  exploring the  root  set  of the  current
activation record. For  any other activation record in  the stack, say
for a function ${\mathit f}$  calling ${\mathit g}$ in the call-chain,
the evaluation  point in ${\mathit f}$  which resulted in  the call to
${\mathit g}$ would be used.

@@

While exploring the heap, the GC  may encounter closures. In RGC mode,
the closure  will be explored  as usual  and copied by  traversing the
references corresponding to the arguments. Since our liveness analysis
predicts the  liveness of  data in its  evaluated form  only, closures
present a challenge for LGC. LGC will  need to map the liveness of the
free variables of the closure to  their evaluation point.  This is not
straight  forward as  variables may  escape their  creation scope  and
needs to  be handled in an  approximate way.  We describe this in detail next.

@@

\section{Experimental Evaluation}

%% \comment{To demonstrate the effectiveness of liveness-based garbage collection,
%% we  have  built a  prototype  consisting  of  an interpreter  for  our
%% language,  a  liveness  analyzer  and  a copying  collector  that  can
%% optionally use the results of liveness analysis for marking instead of
%% reachability. When the collector uses liveness for marking, we call it
%% a  {\em liveness-based  collector} (LGC),  else we  use the  term {\em
%%   reachability-based  collector}  (RGC).   The  copying  collector  is
%% neither incremental nor generational.  As a consequence, any cell that
%% becomes unreachable or  dead is assuredly collected in  the next round
%% of garbage collection.}

@@

\section{Related Work} 
Although augmenting  garbage collection with liveness  information has
been  studied earlier,  it was  mainly  in the  context of  imperative
languages~\cite{Albert,  Hirzel}.  Due  to   the  presence  of  global
variables  and mutation,  practical utility  of simple  liveness based
techniques are found to be inefficient.

>>>>

{\color {Myblue}Section~\ref{sec:defs} introduces
  the syntax and  describes a small-step operational  semantics of the
  programming language  considered in  our analysis.  We  describe our
  liveness     analysis     for      this     lazy     language     in
  Section~\ref{sec:liveness}.   Section~\ref{sec:computing}  describes
  the formulation of this liveness  as a grammar and its approximation
  to DFA. It  also contains a proof showing the  undecidability of the
  precise  liveness in  our formulation.   Section~\ref{sec:GC-scheme}
  discusses different  ways of  handling closures  and their  pros and
  cons.      We     report     our     experimental     results     in
  Section~\ref{sec:experiments} along  with some of  our observations.
  Section~\ref{sec:relatedwork}  discusses  previous work  related  to
  garbage  collection  and liveness  and  Section~\ref{sec:conclusion}
  discusses possible extensions and concludes the paper.  }
\section{The target language---syntax and semantics}
\label{sec:defs}
Figure~\ref{fig:lang-syntax} describes the syntax  of our language. It
is a first order language with lazy semantics. Programs are restricted
to        be        in        Administrative        Normal        Form
(ANF)~\cite{chakravarty03perspective} where  all actual  parameters to
functions  are  variables.  While  this  restriction  does not  affect
expressibility,  this form  has  the benefit  of  making explicit  the
creation of  closures through  the $\LET$  construct.  $\LET$s  in our
language  are lazy;  in the  expression $\LET\,\,  x \leftarrow  s\,\,
\IN\,\, e$, $x$ may occur in  $s$. This enables creation of graph-like
structures in  a pure language. The  restriction of \LET\ to  a single
definition  is for  ease  of  exposition---generalization to  multiple
definitions does not add conceptual difficulties.  We further restrict
each variable in a program to  be distinct, so that no scope shadowing
occurs---this  simplifies  the  reasoning   about  the  program.   The
$\SRETURN$  and   \SIF\  expressions  and  strict   operators  trigger
evaluation of closures.

@@

\section{Liveness}
\label{sec:liveness}

@@

\section{The Garbage Collection Scheme}
\label{sec:GC-scheme}
Our experimental setup consists of  an interpreter for our language, a
liveness  analyzer, and  a single  generation copying  collector.  The
garbage collector  can be  easily configured to  work on the  basis of
reachability (RGC  mode) or use  the liveness DFAs (LGC  mode).  While
exploring the  activation record of a  function, LGC uses  the DFAs at
the latest program  point traversed in the function's  body.  Thus, if
LGC is invoked  at the program point $\pi$, it  would use the liveness
DFAs  at  $\pi$ itself  for  exploring the  root  set  of the  current
activation record. For  any other activation record in  the stack, say
for a function ${\mathit f}$  calling ${\mathit g}$ in the call-chain,
the evaluation  point in ${\mathit f}$  which resulted in  the call to
${\mathit g}$ would be used.

@@

While exploring the heap, the GC  may encounter closures. In RGC mode,
the closure  will be explored  as usual  and copied by  traversing the
references corresponding to the arguments. Since our liveness analysis
predicts the  liveness of  data in its  evaluated form  only, closures
present a challenge for LGC. LGC will  need to map the liveness of the
free variables of the closure to  their evaluation point.  This is not
straight  forward as  variables may  escape their  creation scope  and
need to  be handled in an  approximate way.  We describe this in detail next.

@@

\section{Experimental Evaluation}
\label{sec:experiments}
%% \comment{To demonstrate the effectiveness of liveness-based garbage collection,
%% we  have  built a  prototype  consisting  of  an interpreter  for  our
%% language,  a  liveness  analyzer  and  a copying  collector  that  can
%% optionally use the results of liveness analysis for marking instead of
%% reachability. When the collector uses liveness for marking, we call it
%% a  {\em liveness-based  collector} (LGC),  else we  use the  term {\em
%%   reachability-based  collector}  (RGC).   The  copying  collector  is
%% neither incremental nor generational.  As a consequence, any cell that
%% becomes unreachable or  dead is assuredly collected in  the next round
%% of garbage collection.}

@@

\section{Related Work} 
\label{sec:relatedwork}
Although augmenting  garbage collection with liveness  information has
been  studied earlier,  it was  mainly  in the  context of  imperative
languages~\cite{Albert,  Hirzel}.  Due  to   the  presence  of  global
variables  and mutation,  practical utility  of simple  liveness based
techniques are found to be inefficient.
